Guarded Recursion in Agda via Sized Types

### 2019-02-11 code-oneclock-paper/latex/Introduction.tex
2)
minor changes until page 7

Possibly non-terminating computations
and infinite structures, such as non-wellfounded trees, can be
represented in type theory by extending the type system with
coinductive types. Recursively defined elements of these types
need to be productive in the sense that every
finite unfolding can be computed in a finite number of steps \cite{Coquand93}.
In Agda's encoding of coinductive types using ``musical notation''
\cite{DA10}, productivity is enforced via a strict obligation: in the
definition of a corecursive function, recursive
calls must appear directly under
the application of a constructor. A similar syntactic
check is used in Coq \cite{DBertotK08}.
This restriction
typically makes programming with coinductive types cumbersome,
which spawned the search for alternative techniques to ensure the
well-definedness of corecursive definitions.

@@

Utilizing techniques for representing ``type theory in type theory''
\cite{AltenkirchK16,Chapman09}, we present an Agda formalization of
the syntax of a simple type theory for guarded recursion. This object
language, which we call \GTT, is a variant of Atkey and McBride's type
system for productive coprogramming \cite{atkey2013productive} in
which the clock context can contain at most one clock. The types of
\GTT\ include the aforementioned \IC{▻} modality and a \IC{□} modality,
a nameless analogue of Atkey and McBride's universal clock
quantification. Clouston \etal \cite{CloustonBGB15} studied a
guarded variant of lambda calculus extended with a \IC{□} operation,
which they call ``constant''.  Our object calculus differs from
theirs in that our judgments are indexed by a clock context, which can
be empty or containing exactly one clock. In this respect \GTT\ is
closer in spirit to Atkey and McBride's language. The design of \GTT\
also has the benefit of allowing a more appealing introduction rule for
\IC{□} than Clouston et al.'s.

@@

This paper only include the essential parts of our Agda
formalization. The full code is available at
\url{https://github.com/niccoloveltri/agda-clott}. The formalization
uses Agda 2.5.4.1 and Agda standard library 0.16. The paper is
extracted from a literate Agda file, which implies that all the
displayed code has passed Agda's type and termination checker.

The material is organized in the following way.
In \Cref{sec:prelim}, we discuss the metatheory and we give an
overview of sized types.
In \Cref{sec:syntax}, we introduce the syntax of \GTT, our object type theory for guarded recursion.
Subsequently, we give a categorical semantics.
In \Cref{sec:presheaf_sem}, we show how to implement presheaves over
sizes and how to model the fragment of
\GTT\ corresponding to simply typed lambda calculus.
In \Cref{sec:guarded}, we discuss the semantics of the guarded
recursive and coinductive features.
In \Cref{sec:consistent}, we describe how to formally interpret the
syntax into the model, which entails the consistency of the syntax.
Finally, in \Cref{sec:concl}, we draw conclusions and suggest  future
directions of work.

>>>>

Possibly non-terminating computations
and infinite structures, such as non-wellfounded trees, can be
represented in type theory by extending the type system with
coinductive types. Recursively defined elements of these types
need to be productive in the sense that every
finite unfolding can be computed in a finite number of steps \cite{Coquand93}.
In Agda's encoding of coinductive types using ``musical notation''
\cite{DA10}, productivity is enforced via a strict obligation: in the
definition of a corecursive function, recursive
calls must appear directly under
the application of a constructor. A similar syntactic
check is used in Coq. % \cite{DBertotK08}.
This restriction
typically makes programming with coinductive types cumbersome,
which spawned the search for alternative techniques to ensure the
well-definedness of corecursive definitions.

@@

Utilizing techniques for representing ``type theory in type theory''
\cite{AltenkirchK16,Chapman09}, we present an Agda formalization of
the syntax of a simple type theory for guarded recursion. This object
language, which we call \GTT, is a variant of Atkey and McBride's type
system for productive coprogramming \cite{atkey2013productive} in
which the clock context can contain at most one clock. The types of
\GTT\ include the aforementioned \IC{▻} modality and a \IC{□} modality,
a nameless analogue of Atkey and McBride's universal clock
quantification. Clouston \etal \cite{CloustonBGB15} studied a
guarded variant of lambda calculus extended with a \IC{□} operation,
which they call ``constant''.  Our object calculus differs from
theirs in that our judgments are indexed by a clock context, which can
be empty or containing exactly one clock. 
%In this respect \GTT\ is closer in spirit to Atkey and McBride's
%language.
The design of \GTT\ has the benefit of allowing a more appealing introduction rule for
\IC{□} than Clouston et al.'s.

@@

This paper only include the essential parts of our Agda
formalization. The full code is available at
\url{https://github.com/niccoloveltri/agda-clott}. The formalization
uses Agda 2.5.4.2 and Agda standard library 0.16. The paper is
extracted from a literate Agda file, which implies that all the
displayed code has passed Agda's type and termination checker.

The material is organized in the following way.
In \Cref{sec:prelim}, we discuss the metatheory and we give an
overview of sized types.
In \Cref{sec:syntax}, we introduce the syntax of \GTT, our object type theory for guarded recursion.
Subsequently, we give a categorical semantics.
In \Cref{sec:presheaf_sem}, we show how to implement presheaves over
sizes and how to model the fragment of
\GTT\ corresponding to simply typed lambda calculus.
In \Cref{sec:guarded}, we discuss the semantics of the guarded
recursive and coinductive features.
In \Cref{sec:consistent}, we prove the object language sound \wrt the
categorical model,
%describe how to formally interpret the syntax into the model, 
which in turn entails the consistency of the syntax.
Finally, in \Cref{sec:concl}, we draw conclusions and suggest  future
directions of work.
